---
output:
  pdf_document: default
  html_document: default
---
%All you need to do is knit this to pdf and it will render the tex file. As always, let me know if something isnt working or rendering correctly. I used AI to OCR and format my answers, and to brainstorm with me how to answer the lab questions, which I then used to help me answer the Pset. In this case, all errors are my own, and I am more than happy to provide my written work as well. 

---
title: "Applied Statistical Programming: Problem Set 2"
author: "Bowen Damask"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    highlight: tango
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \renewcommand{\labelenumi}{\textbf{Question \arabic{enumi}.}}
  - \renewcommand{\labelenumii}{\textbf{(\alph{enumii})}}
  - \allowdisplaybreaks
  - \AtBeginDocument{\setlength{\abovedisplayskip}{6pt}\setlength{\belowdisplayskip}{6pt}\setlength{\abovedisplayshortskip}{3pt}\setlength{\belowdisplayshortskip}{3pt}}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4)
```

```{=latex}
\begin{enumerate}
```

\item \leavevmode

*Imagine that $(y_1, y_2, y_3, y_4, y_5) \sim \mathrm{Mult}(n, \delta_1, \delta_2, \delta_3, \delta_4, \delta_5)$, with $\delta_1 = \frac{1}{4}\theta$, $\delta_2 = \delta_3 = \frac{1}{4}(1-\theta)$, $\delta_4 = \frac{1}{2}$, and $\delta_5 = \frac{1}{4}\theta$. Instead of observing these data, we observe $(x_1, x_2, x_3, x_4) \sim \mathrm{Mult}(n, \pi_1, \pi_2, \pi_3, \pi_4)$, where $\pi_1 = \frac{1}{4}\theta$, $\pi_2 = \pi_3 = \frac{1}{4}(1-\theta)$, and $\pi_4 = \frac{1}{2} + \frac{1}{4}\theta$. So we have that $x_4 = y_4 + y_5$. Use the EM algorithm to recover an estimate of $\theta$.*

Likelihood is just written as

\begin{align*}
L_c(\theta) &= \frac{n!}{\prod_{k} y_k!} \left( \tfrac{\theta}{4} \right)^{y_1} \left( \tfrac{1-\theta}{4} \right)^{y_2} \left( \tfrac{1-\theta}{4} \right)^{y_3} \left( \tfrac{1}{2} \right)^{y_4} \left( \tfrac{\theta}{4} \right)^{y_5} \\
\ell_c(\theta) &= y_1 \log \tfrac{\theta}{4} + y_2 \log \tfrac{1-\theta}{4} + y_3 \log \tfrac{1-\theta}{4} + y_4 \log \tfrac{1}{2} + y_5 \log \tfrac{\theta}{4} + \log \frac{n!}{\prod_k y_k!} \\
&= (y_1 + y_5) \log \theta + (y_2 + y_3) \log (1 - \theta) + C
\end{align*}

Taking expected value of $y_5$ we get

\begin{align*}
y_5 \mid x_4, \theta &\sim \mathrm{Bin}\!\left( x_4, \; \frac{\delta_5}{\delta_4 + \delta_5} \right) \\
\frac{\delta_5}{\delta_4 + \delta_5} &= \frac{\theta/4}{1/2 + \theta/4} \\
&= \frac{\theta}{2 + \theta} \\
y_5^{(t)} &\equiv E\!\left[ y_5 \mid x, \theta^{(t)} \right] \\
&= x_4 \, \frac{\theta^{(t)}}{2 + \theta^{(t)}}
\end{align*}

Taking expectations we get

\begin{align*}
Q(\theta \mid \theta^{(t)}) &= E\!\left[ \ell_c(\theta) \mid x, \theta^{(t)} \right] \\
&= \left( E[y_1 \mid x] + E[y_5 \mid x, \theta^{(t)}] \right) \log \theta + E[y_2 + y_3 \mid x] \log (1 - \theta) + C \\
&= \left( x_1 + y_5^{(t)} \right) \log \theta + (x_2 + x_3) \log (1 - \theta) + C
\end{align*}

Then just max $Q$ over $\theta$ to get

\begin{align*}
\frac{\partial Q}{\partial \theta} &= \frac{x_1 + y_5^{(t)}}{\theta} - \frac{x_2 + x_3}{1 - \theta} = 0 \\
\left( x_1 + y_5^{(t)} \right)(1 - \theta) &= (x_2 + x_3) \theta \\
x_1 + y_5^{(t)} &= \left( x_1 + y_5^{(t)} + x_2 + x_3 \right) \theta \\
\theta^{(t+1)} &= \frac{x_1 + y_5^{(t)}}{x_1 + y_5^{(t)} + x_2 + x_3}
\end{align*}

and repeat until

\begin{align*}
\left| \theta^{(t+1)} - \theta^{(t)} \right| &< \epsilon
\end{align*}

\item \leavevmode

*Let $M_{ij}$ indicate whether records $i$ and $j$ are a match. We do not observe $M_{ij}$, but we do observe their agreement on $K$ variables. Define $\delta_k(i,j) = 1$ if the records agree on variable $k$, and 0 otherwise. Each pair has an observed comparison vector*
\begin{align*}
\boldsymbol{\delta}(i,j) = \big( \delta_1(i,j), \dots, \delta_K(i,j) \big)
\end{align*}
*For $n$ record pairs, use the EM algorithm to estimate the probability that each pair is a match, along with the model parameters. Assume that conditional on $M_{ij}$ each component of $\boldsymbol{\delta}(i,j)$ is independent of each other.*

*Hint: given the information above, if you make the right calls for the distributions of $M_{ij}$ and of $\boldsymbol{\delta}(i,j) \mid M_{ij} = m$, where $m \in \{0, 1\}$, then you will be good to go.*

Define \(\delta_{ij} \equiv \boldsymbol{\delta}(i,j)\) and
\(\psi = (\lambda, m_1, \dots, m_K, u_1, \dots, u_K)\). Then
\begin{align*}
M_{ij} &\sim \mathrm{Bernoulli}(\lambda) \\
f_1(\delta_{ij}) \equiv \Pr(\delta_{ij} \mid M_{ij} = 1) &= \prod_{k=1}^{K} m_k^{\delta_k(i,j)} (1 - m_k)^{1 - \delta_k(i,j)} \\
f_0(\delta_{ij}) \equiv \Pr(\delta_{ij} \mid M_{ij} = 0) &= \prod_{k=1}^{K} u_k^{\delta_k(i,j)} (1 - u_k)^{1 - \delta_k(i,j)}
\end{align*}

Then we can just get the observed likelihood as

\begin{align*}
\Pr(\delta_{ij}) &= \Pr(M_{ij} = 1) \Pr(\delta_{ij} \mid M_{ij} = 1) + \Pr(M_{ij} = 0) \Pr(\delta_{ij} \mid M_{ij} = 0) \\
&= \lambda f_1(\delta_{ij}) + (1 - \lambda) f_0(\delta_{ij}) \\
L(\psi) &= \prod_{(i,j)} \Big[ \lambda f_1(\delta_{ij}) + (1 - \lambda) f_0(\delta_{ij}) \Big] \\
\ell(\psi) &= \sum_{(i,j)} \log \Big[ \lambda f_1(\delta_{ij}) + (1 - \lambda) f_0(\delta_{ij}) \Big]
\end{align*}

If we observed $M_{ij}$, the complete likelihood would be

\begin{align*}
\Pr(\delta_{ij}, M_{ij}) &= \big[ \lambda f_1(\delta_{ij}) \big]^{M_{ij}} \big[ (1 - \lambda) f_0(\delta_{ij}) \big]^{1 - M_{ij}} \\
L_c(\psi) &= \prod_{(i,j)} \big[ \lambda f_1(\delta_{ij}) \big]^{M_{ij}} \big[ (1 - \lambda) f_0(\delta_{ij}) \big]^{1 - M_{ij}} \\
\ell_c(\psi) &= \sum_{(i,j)} \Big\{ M_{ij} \big[ \log \lambda + \log f_1(\delta_{ij}) \big] + (1 - M_{ij}) \big[ \log (1 - \lambda) + \log f_0(\delta_{ij}) \big] \Big\} \\
\log f_1(\delta_{ij}) &= \sum_{k=1}^{K} \big[ \delta_k(i,j) \log m_k + (1 - \delta_k(i,j)) \log (1 - m_k) \big] \\
\log f_0(\delta_{ij}) &= \sum_{k=1}^{K} \big[ \delta_k(i,j) \log u_k + (1 - \delta_k(i,j)) \log (1 - u_k) \big]
\end{align*}

Taking expectation we can get posterior pr(x)

\begin{align*}
\xi_{ij}^{(t)} &\equiv E\!\left[ M_{ij} \mid \delta_{ij}, \psi^{(t)} \right] \\
&= \Pr\!\left( M_{ij} = 1 \mid \delta_{ij}, \psi^{(t)} \right) \\
&= \frac{\Pr(M_{ij} = 1) \Pr(\delta_{ij} \mid M_{ij} = 1)}{\Pr(\delta_{ij})} \\
&= \frac{\lambda^{(t)} f_1^{(t)}(\delta_{ij})}{\lambda^{(t)} f_1^{(t)}(\delta_{ij}) + (1 - \lambda^{(t)}) f_0^{(t)}(\delta_{ij})}
\end{align*}

Taking expectations we get

\begin{align*}
Q(\psi \mid \psi^{(t)}) &= E\!\left[ \ell_c(\psi) \mid \delta, \psi^{(t)} \right] \\
&= \sum_{(i,j)} \Big\{ \xi_{ij}^{(t)} \big[ \log \lambda + \log f_1(\delta_{ij}) \big] + (1 - \xi_{ij}^{(t)}) \big[ \log (1 - \lambda) + \log f_0(\delta_{ij}) \big] \Big\}
\end{align*}

We can then update for $\lambda$

\begin{align*}
\frac{\partial Q}{\partial \lambda} &= \frac{\sum_{(i,j)} \xi_{ij}^{(t)}}{\lambda} - \frac{\sum_{(i,j)} (1 - \xi_{ij}^{(t)})}{1 - \lambda} = 0 \\
(1 - \lambda) \sum_{(i,j)} \xi_{ij}^{(t)} &= \lambda \sum_{(i,j)} (1 - \xi_{ij}^{(t)}) \\
\sum_{(i,j)} \xi_{ij}^{(t)} - \lambda \sum_{(i,j)} \xi_{ij}^{(t)} &= \lambda n - \lambda \sum_{(i,j)} \xi_{ij}^{(t)} \\
\sum_{(i,j)} \xi_{ij}^{(t)} &= \lambda n \\
\lambda^{(t+1)} &= \frac{1}{n} \sum_{(i,j)} \xi_{ij}^{(t)}
\end{align*}

and for $m_k$

\begin{align*}
\frac{\partial Q}{\partial m_k} &= \sum_{(i,j)} \xi_{ij}^{(t)} \left[ \frac{\delta_k(i,j)}{m_k} - \frac{1 - \delta_k(i,j)}{1 - m_k} \right] = 0 \\
(1 - m_k) \sum_{(i,j)} \xi_{ij}^{(t)} \delta_k(i,j) &= m_k \sum_{(i,j)} \xi_{ij}^{(t)} (1 - \delta_k(i,j)) \\
\sum_{(i,j)} \xi_{ij}^{(t)} \delta_k(i,j) &= m_k \sum_{(i,j)} \xi_{ij}^{(t)} \\
m_k^{(t+1)} &= \frac{\sum_{(i,j)} \xi_{ij}^{(t)} \delta_k(i,j)}{\sum_{(i,j)} \xi_{ij}^{(t)}}
\end{align*}

and also for $u_k$

\begin{align*}
\frac{\partial Q}{\partial u_k} &= \sum_{(i,j)} (1 - \xi_{ij}^{(t)}) \left[ \frac{\delta_k(i,j)}{u_k} - \frac{1 - \delta_k(i,j)}{1 - u_k} \right] = 0 \\
(1 - u_k) \sum_{(i,j)} (1 - \xi_{ij}^{(t)}) \delta_k(i,j) &= u_k \sum_{(i,j)} (1 - \xi_{ij}^{(t)}) (1 - \delta_k(i,j)) \\
\sum_{(i,j)} (1 - \xi_{ij}^{(t)}) \delta_k(i,j) &= u_k \sum_{(i,j)} (1 - \xi_{ij}^{(t)}) \\
u_k^{(t+1)} &= \frac{\sum_{(i,j)} (1 - \xi_{ij}^{(t)}) \delta_k(i,j)}{\sum_{(i,j)} (1 - \xi_{ij}^{(t)})}
\end{align*}

We repeat until the log-likelihood converges, which gives

\begin{align*}
\left| \ell(\psi^{(t+1)}) - \ell(\psi^{(t)}) \right| &< \epsilon \\
\widehat{\Pr}(M_{ij} = 1 \mid \delta_{ij}) &= \hat{\xi}_{ij} \\
&= \frac{\hat{\lambda} \hat{f}_1(\delta_{ij})}{\hat{\lambda} \hat{f}_1(\delta_{ij}) + (1 - \hat{\lambda}) \hat{f}_0(\delta_{ij})}
\end{align*}

\pagebreak[3]
\item \leavevmode

*Consider the following table of counts for the agreement patterns $\boldsymbol{\delta}$ defined in the previous question.* *Here, $\boldsymbol{\delta} = (0, 0, 0)$ indicates disagreement on all four fields and occurs for 172738 record pairs. Similarly, $\boldsymbol{\delta} = (1, 1, 1)$ indicates agreement on all four fields and occurs for 50 record pairs. Patterns omitted from the table have a count of zero.*

*These counts are the observed data; as in the previous question, the match status $M$ remains unobserved for every pair. Using your derivations from the previous question, implement the EM algorithm and fit the model to these counts.*

*Store the observed-data log-likelihood at each iteration. Plot it against the iteration number and verify that it does not decrease.*

```{r, fig.width = 6.2, fig.height = 2.6}
set.seed(21523514)
delta <- rbind(c(0, 0, 0), c(1, 0, 0), c(0, 1, 0), c(0, 0, 1), c(1, 1, 1))
colnames(delta) <- c("d1", "d2", "d3")
w <- c(172738, 2030, 114, 68, 50)

## pr(x) of each pattern is just
pattern_prob <- function(p) {
  apply(delta, 1, function(d) prod(ifelse(d == 1, p, 1 - p)))
}

lambda <- 0.01
m      <- rep(0.9, 3)
u      <- rep(0.1, 3)
loglik <- c()

for (t in 1:1000) {
  ## Pr(pattern pair is a match)
  match    <- lambda       * pattern_prob(m)
  nonmatch <- (1 - lambda) * pattern_prob(u)
  xi       <- match / (match + nonmatch)

  ## observed-data log-likelihood at the current parameters
  loglik[t] <- sum(w * log(match + nonmatch))

  ## E(number of matches/non-matches)
  n_match    <- w * xi
  n_nonmatch <- w * (1 - xi)

  lambda <- sum(n_match) / sum(w)
  m      <- colSums(n_match * delta)    / sum(n_match)
  u      <- colSums(n_nonmatch * delta) / sum(n_nonmatch)

  ## stop when the log-likelihood stops changing
  if (t > 1 && abs(loglik[t] - loglik[t - 1]) < 1e-8) break
}

lambda
round(rbind(m, u), 6)

## estimated match pr(x) 
data.frame(delta, count = w, Pr_match = round(xi, 4))

## plotted
plot(loglik, type = "b", xlab = "Iteration", ylab = "Log-likelihood")

## never decreases?
all(diff(loglik) >= -1e-8)
```

\item \leavevmode

\emph{Suppose \(X_i\) with \(i = \{1, \dots, n\}\), comes from a
two-component Gaussian mixture. For \(n_1\) observations, the class
label \(Z_i\) is known; for the rest i.e., \(n_2 = n - n_1\), the class
labels are missing. Derive the EM algorithm for this problem.}

Start by letting \(\mathcal{L} = \{1, \dots, n_1\}\),
\(\mathcal{U} = \{n_1 + 1, \dots, n\}\),
\(z_{ic} = \mathbb{1}\{Z_i = c\}\) and
\(\psi = (\pi_1, \mu_1, \mu_2, \sigma_1^2, \sigma_2^2)\) 
\begin{align*}
\Pr(Z_i = c) &= \pi_c, \qquad c \in \{1, 2\}, \qquad \pi_1 + \pi_2 = 1 \\
\phi_c(x_i) \equiv f(x_i \mid Z_i = c) &= \frac{1}{\sqrt{2\pi \sigma_c^2}} \exp\!\left( -\frac{(x_i - \mu_c)^2}{2\sigma_c^2} \right)
\end{align*}

Then the complete likelihood can just be written as 
\begin{align*}
L_c(\psi) &= \prod_{i=1}^{n} \prod_{c=1}^{2} \big[ \pi_c \phi_c(x_i) \big]^{z_{ic}} \\
\ell_c(\psi) &= \sum_{i=1}^{n} \sum_{c=1}^{2} z_{ic} \Big[ \log \pi_c + \log \phi_c(x_i) \Big] \\
&= \sum_{i=1}^{n} \sum_{c=1}^{2} z_{ic} \Big[ \log \pi_c - \tfrac{1}{2} \log (2\pi) - \tfrac{1}{2} \log \sigma_c^2 - \frac{(x_i - \mu_c)^2}{2\sigma_c^2} \Big]
\end{align*}

Taking the posterior pr(x) we get 
\begin{align*}
\tau_{ic}^{(t)} &\equiv E\!\left[ z_{ic} \mid x, \{z_j\}_{j \in \mathcal{L}}, \psi^{(t)} \right] \\
\tau_{ic}^{(t)} &= z_{ic}, \qquad i \in \mathcal{L} \\
\tau_{ic}^{(t)} &= \Pr(Z_i = c \mid x_i, \psi^{(t)}), \qquad i \in \mathcal{U} \\
&= \frac{\Pr(Z_i = c) f(x_i \mid Z_i = c)}{f(x_i)} \\
&= \frac{\pi_c^{(t)} \phi_c^{(t)}(x_i)}{\pi_1^{(t)} \phi_1^{(t)}(x_i) + \pi_2^{(t)} \phi_2^{(t)}(x_i)}
\end{align*}

Then the expected complete log-likelihood is  
\begin{align*}
Q(\psi \mid \psi^{(t)}) &= E\!\left[ \ell_c(\psi) \mid x, \{z_j\}_{j \in \mathcal{L}}, \psi^{(t)} \right] \\
&= \sum_{i=1}^{n} \sum_{c=1}^{2} \tau_{ic}^{(t)} \Big[ \log \pi_c - \tfrac{1}{2} \log (2\pi) - \tfrac{1}{2} \log \sigma_c^2 - \frac{(x_i - \mu_c)^2}{2\sigma_c^2} \Big]
\end{align*}

And the expected class counts are 
\begin{align*}
N_c^{(t)} &\equiv \sum_{i=1}^{n} \tau_{ic}^{(t)} \\
&= \sum_{i \in \mathcal{L}} z_{ic} + \sum_{i \in \mathcal{U}} \tau_{ic}^{(t)} \\
N_1^{(t)} + N_2^{(t)} &= n
\end{align*}

Finally, we update for \(\pi_1\) 
\begin{align*}
Q &= N_1^{(t)} \log \pi_1 + N_2^{(t)} \log (1 - \pi_1) + C \\
\frac{\partial Q}{\partial \pi_1} &= \frac{N_1^{(t)}}{\pi_1} - \frac{N_2^{(t)}}{1 - \pi_1} = 0 \\
(1 - \pi_1) N_1^{(t)} &= \pi_1 N_2^{(t)} \\
N_1^{(t)} &= \pi_1 \left( N_1^{(t)} + N_2^{(t)} \right) \\
N_1^{(t)} &= \pi_1 n \\
\pi_1^{(t+1)} &= \frac{N_1^{(t)}}{n} \\
\pi_2^{(t+1)} &= 1 - \pi_1^{(t+1)}
\end{align*}

Then we update for \(\mu_c\) 
\begin{align*}
\frac{\partial Q}{\partial \mu_c} &= \sum_{i=1}^{n} \tau_{ic}^{(t)} \frac{x_i - \mu_c}{\sigma_c^2} = 0 \\
\sum_{i=1}^{n} \tau_{ic}^{(t)} (x_i - \mu_c) &= 0 \\
\sum_{i=1}^{n} \tau_{ic}^{(t)} x_i &= \mu_c \sum_{i=1}^{n} \tau_{ic}^{(t)} \\
\mu_c^{(t+1)} &= \frac{\sum_{i=1}^{n} \tau_{ic}^{(t)} x_i}{N_c^{(t)}}
\end{align*}

and also for \(\sigma_c^2\) 
\begin{align*}
\frac{\partial Q}{\partial \sigma_c^2} &= \sum_{i=1}^{n} \tau_{ic}^{(t)} \left[ -\frac{1}{2\sigma_c^2} + \frac{(x_i - \mu_c)^2}{2\sigma_c^4} \right] = 0 \\
\sum_{i=1}^{n} \tau_{ic}^{(t)} \left[ -\sigma_c^2 + (x_i - \mu_c)^2 \right] &= 0 \\
\sigma_c^2 \sum_{i=1}^{n} \tau_{ic}^{(t)} &= \sum_{i=1}^{n} \tau_{ic}^{(t)} (x_i - \mu_c)^2 \\
\sigma_c^{2\,(t+1)} &= \frac{\sum_{i=1}^{n} \tau_{ic}^{(t)} \left( x_i - \mu_c^{(t+1)} \right)^2}{N_c^{(t)}}
\end{align*}

and repeat until
\begin{align*}
\left\| \psi^{(t+1)} - \psi^{(t)} \right\| &< \epsilon
\end{align*}

```{=latex}
\end{enumerate}
```
